### 2018-01-10 ident/docs/ident_manuscript.tex
improve methods + strat with results

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{tabu}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage[space]{grffile}
\usepackage{subfigure}
\usepackage{longtable}
\usepackage{multirow}
\usepackage[margin=1.0in]{geometry}
\renewcommand{\baselinestretch}{2.0}

@@

\subsection{Kinetic model of gluconeogenesis in E. coli:}\label{sec:small-model}
The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
\begin{equation}\label{eq:ode1}
\frac{d}{dt}pep=v_1-v_2-v_4
\end{equation}
\begin{equation}\label{eq:ode2}
\frac{d}{dt}fdp=v_2-v_3
\end{equation}
\begin{equation}\label{eq:ode3}
\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
\end{equation}
The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
\begin{equation}\label{eq:flux1}
v_1 = k_{1}^{cat}E\frac{acetate}{acetate+K_{1}^{acetate}}
\end{equation}	
\begin{equation}\label{eq:flux2}
v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
\end{equation}
\begin{equation}\label{eq:flux3}
v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
\end{equation}
The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
\begin{equation}\label{eq:flux4}
v_4 = k_{4}^{cat}.pep
\end{equation}

\subsection{Data for establishing parameter identifiability in kinetic model of gluconeogenesis:}
Steady state metabolomics and fluxomics data can be gathered under different physiological conditions by either perturbing the expression levels for different enzymes within a metabolic network, or by changing the substrate concentrations under which the cells grow. The aforementioned model of gluconeogenesis has three different fluxes ($v_1$, $v_2$ and $v_3$) whose enzyme expression parameters ($V_1^{max}$, $V_2^{max}$ and $V_3^{max}$) can be perturbed to simulate the repression and over expression of the corresponding enzymes. Furthermore, the acetate concentration, that determines the acetate uptake flux $v_1$, can also be perturbed to measure cellular response to changes in the substrate concentration. We use the in silico data generated by 18 different experiments wherein these four model parameters ($acetate$, $V_1^{max}$, $V_2^{max}$ and $V_3^{max}$) were perturbed to demonstrate practical parameter identification with our methodology. The experiments and the perturbed values of each of the four the parameters are given in the Appendix.

\subsection{Identifiability of parameters in a kinetic model of gluconeogenesis:}\label{sec:example}
Here, we demonstrate the use of our computer algebra-based methodology to establish practical identifiability of parameters for flux $v_2$ in the small model of gluconeogenesis described in Section \ref{sec:small-model}. We choose flux $v_2$, expressed using Michaelis-Menten kinetics for this demonstration. $v_2$ has two parameters, $V_2^{max}$ and $K_2^{pep}$ that need to be identifiable so that they can be estimated from experimental data. Here, we assume that data (for the concentrations and fluxes) from at least two different sets of experiments is available. We label the available concentrations as ${pep}^1$ and ${pep}^2$, and the fluxes as ${v_2}^1$ and ${v_2}^2$ respectively, from each experiment. Accordingly, the nonlinear algebraic equations shown in Equation (\ref{eq:nonlineq}) can be formulated for flux $v_2$ as follows:
\begin{subequations}\label{eq:nonlin-flux2}
	\begin{align}
	{v_2}^1 = V_{2}^{max}\frac{pep^1}{pep^1+K_{2}^{pep}}\\
	{v_2}^2 = V_{2}^{max}\frac{pep^2}{pep^2+K_{2}^{pep}}
	\end{align}
\end{subequations}
Solving this simultaneous system of equations using Mathematica (Wolfram Research, USA), a computer algebra system, we get two nonlinear algebraic equations in the parameters $V_2^{max}$ and $K_2^{pep}$ based on the form shown earlier in Equation (\ref{eq:theta-eq}).
\begin{subequations}\label{eq:v2_par}
	\begin{align}		
	V_2^{max} = \frac{v_2^1v_2^2(pep^1-pep^2)}{v_2^2pep^1-v_2^1pep^2}\\
	K_2^{pep} = \frac{pep^1(v_2^1pep^2-v_2^2pep^2)}{v_2^2pep^1-v_2^1pep^2}
	\end{align}
\end{subequations}
In Equation (\ref{eq:v2_par}), the denominator of the right hand side expression is used to test the identifiability of parameters $V_2^{max}$ (Equation \ref{eq:v2_par}a) and $K_2^{pep}$ (Equation \ref{eq:v2_par}b) for different available experimental data set combinations. Since the enzyme binding constant ($K_2^{pep}$) and maximum reaction rate ($V_2^{max}$) cannot be negative, we can further reduce the criteria for identifiability for both these parameters by saying that the evaluated expressions should be non-negative.

\section{Results:}
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=.6\textwidth,height=.6\textheight,keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure3/ident_analysis}}
	\caption{to be filled in}\label{fig:figure1}
\end{figure}

@@

\section{Results:}
First, in Section \ref{sec:example} we demonstrate how our methodology can be applied for one of the fluxes in the gluconeogenic model. In Section \ref{sec:closedform} that follows, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.

\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
The complexity of the equations in our specific scenario is determined by the complexity of enzyme-metabolite interaction models used describe fluxes in metabolic networks. Although computer algebra systems (CAS) are capable of handling complex symbolic calculations, sometimes, the complexity of getting closed form expressions for all kinetic parameters of certain rate law formulations is too much for the CAS to overcome. We encountered this scenario in the case of the gluconeogenic model for flux $v_3$ where the kinetics of the allosterically activated reaction are described by the MWC model for allosteric regulation. In order to overcome this computational difficulty, we used a convenience kinetic rate law formulation to describe the allosteric interaction in $v_3$. We give this formulation in Section.

We believe this problem of computational tractability to occur in other complex kinetic rate law formulations as well. Examples?

\subsection{Combinations of experiments that will enable identification of all model parameters:}
\subsection{Expanding methodology to RBC model:}

If any data generated from a perturbation experiment $i$ results in nonidentifiability, we eliminate experiment $i$ from the list of experiments that need to be performed for parameter estimation.

\begin{table}[!tbhp]
	\caption{Table showing the perturbed values of all fluxes used for parameter estimation.}
	\begin{center}				
		\begin{tabular}{ccc}
			Designation & Perturbed Fluxes & Perturbed Values\\
			\hline
			P1 & $v_1$ & 2\\
			P2 & $v_2$ & 0.2\\
			P3 & $v_3$ & 0.5
		\end{tabular}
	\end{center}	
	\label{tab:pval}
\end{table}

\paragraph{Outline:}
\begin{itemize}
	\item parameter estimation is a well developed field typically using minimization of least square error to estimate model parameters from available experimental data
	\item if parameters are structurally identifiable, it does not guarantee practical identifiability from noisy experimental data
	\item identifiability dependent on whether given datasets (outputs) for estimation can sufficiently distinguish between different parameter values
\end{itemize}

\paragraph{Sections:}
\begin{itemize}
	\item datasets required for parameter estimation in kinetic models of metabolism (methods?)		
	\item identifiability in kotte model - scalability, number of experiments required, requirements for time course data(? in the intro)		
	\item identifiability in large rbc model
\end{itemize}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tabu}
\usepackage{lmodern}
\usepackage{graphicx}
\usepackage[space]{grffile}
\usepackage{subfigure}
\usepackage{longtable}
\usepackage{multirow}
\usepackage[margin=1.0in]{geometry}
\renewcommand{\baselinestretch}{2.0}

@@

\subsection{Kinetic model of gluconeogenesis in E. coli:}\label{sec:small-model}
The proposed model for acetate consumption through gluconeogenesis and its corresponding kinetic model is used as a case study to illustrate the utility of identifiability analysis for the design of experiments for estimating parameters in kinetic models of metabolism. The kinetic model is described below.
\begin{equation}\label{eq:ode1}
\frac{d}{dt}pep=v_1-v_2-v_4
\end{equation}
\begin{equation}\label{eq:ode2}
\frac{d}{dt}fdp=v_2-v_3
\end{equation}
\begin{equation}\label{eq:ode3}
\frac{d}{dt}E=v_{e,max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right) - d E
\end{equation}
The kinetic expressions for fluxes $v_1$ through $v_4$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
\begin{equation}\label{eq:flux1}
v_1 = k_{1}^{cat}E\frac{ac}{ac+K_{1}^{ac}}
\end{equation}	
\begin{equation}\label{eq:flux2}
v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
\end{equation}
\begin{equation}\label{eq:flux3}
v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
\end{equation}
The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$. The added flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
\begin{equation}\label{eq:flux4}
v_4 = k_{4}^{cat}.pep
\end{equation}

\subsection{Identifiability of parameters in a kinetic model of gluconeogenesis:}\label{sec:example}
Here, we demonstrate the use of our computer algebra-based methodology to establish practical identifiability of parameters for fluxes $v_1$, $v_2$ and $v_3$ in the small model of gluconeogenesis described in Section \ref{sec:small-model}. For the purposes of our identifiability methodology, we assume that relevant steady state metabolite concentrations and fluxes can be measured.

In flux $v_1$, the concentration of the enzyme E is used as a variable. If the enzyme concentration can be measured, then the expression for $v_1$ given in Equation (\ref{eq:flux1}) can be used for identifiability analysis of parameters $k_1^{cat}$ and $K_1^{ac}$. However, if enzyme concentrations are not available, the expression in Equation (\ref{eq:flux1}) can be modified as given in Equation (\ref{eq:flux1a}) below. This equation does not make use of enzyme concentrations as variables, and uses $V_1^{max}$ and $K_1^{ac}$ as parameters.
\begin{align}\label{eq:flux1a}
v_1 = V_1^{max}\frac{ac}{ac+K_{1}^{ac}}
\end{align}

We choose this expression for flux $v_1$ (Equation \ref{eq:flux1a}), expressed using Michaelis-Menten kinetics, to demonstrate our method for practical identifiability. Both $V_1^{max}$ and $K_1^{ac}$ need to be identifiable so that they can be estimated from experimental data. Here, we assume that data (for the concentrations and fluxes) from at least two different sets of experiments is available i.e., in Equation (\ref{eq:nonlineq}) $k = {1, 2}$ . We label the available concentrations and fluxes as ${ac}^{(k)}$ and ${v_1}^{(k)}$, respectively.  Accordingly, the nonlinear algebraic equations shown in Equation (\ref{eq:nonlineq}) can be formulated for $v_1$ as follows:
\begin{align}\label{eq:nonlin-flux2}
{v_1}^{(k)} = V_{1}^{max}\frac{ac^{(k)}}{ac^{(k)}+K_{1}^{ac}} && \forall k=\{1, 2\}
\end{align}

Solving this simultaneous system of $k$ equations using Mathematica (Wolfram Research, USA), a computer algebra system, we get $p=2$ nonlinear algebraic equations in the parameters $V_1^{max}$ and $K_1^{ac}$ based on the form shown earlier in Equation (\ref{eq:theta-eq}).
\begin{subequations}\label{eq:v1_par}
	\begin{align}		
	V_1^{max} = \frac{v_1^{(1)}v_1^{(2)}(ac^{(1)}-ac^{(2)})}{v_1^{(2)}ac^{(1)}-v_1^{(1)}ac^{(2)}}\\
	K_1^{ac} = \frac{ac^{(1)}ac^{(2)}(v_1^{(1)}-v_1^{(2)})}{v_1^{(2)}ac^{(1)}-v_1^{(1)}ac^{(2)}}
	\end{align}
\end{subequations}
In Equation (\ref{eq:v1_par}), the denominator of the right hand side expression is used to test the identifiability of parameters $V_1^{max}$ (Equation \ref{eq:v1_par}a) and $K_1^{ac}$ (Equation \ref{eq:v1_par}b) for different available experimental data set combinations. Since the enzyme binding constant ($K_1^{ac}$) and maximum reaction rate ($V_1^{max}$) cannot be negative, we can further reduce the criteria for identifiability for both these parameters by saying that the evaluated expressions should be non-negative.

\subsection{Data for establishing parameter identifiability in kinetic model of gluconeogenesis:}
Steady state metabolomics and fluxomics data can be gathered under different physiological conditions by either perturbing the expression levels for different enzymes within a metabolic network, or by changing the substrate concentrations under which the cells grow. The aforementioned model of gluconeogenesis has three different fluxes ($v_1$, $v_2$ and $v_3$) whose enzyme expression parameters ($V_1^{max}$, $V_2^{max}$ and $V_3^{max}$) can be perturbed to simulate the repression and over expression of the corresponding enzymes. Furthermore, the acetate concentration, that determines the acetate uptake flux $v_1$, can also be perturbed to measure cellular response to changes in the substrate concentration. We use the in silico data generated by 18 different experiments wherein these four model parameters ($ac$, $V_1^{max}$, $V_2^{max}$ and $V_3^{max}$) are perturbed to demonstrate practical parameter identification with our methodology. The experiments and the perturbed values of each of the four the parameters are given in the Appendix.

The minimum number of experiments from which data is required for identifying all the parameters of a given flux is determined by dimension $\mathbb{R}^p$ of the parameter space of a chosen flux $v_i$. For instance, as demonstrated above, data from two distinct experiments is required for identifying the two parameters of $v_1$. In this case $p=2$. This also applies for identifying parameters in $v_2$.

Looking at all the three fluxes whose parameters have to be identified, $v_3$ has the maximum number of parameters at three. Hence, we choose to select multiple distinct data sets, each with three different experiments to establish the practical identifiability of all the parameters in the model. Accordingly, we can select 4896 (18 x 17 x 16) distinct combinations of three experiments from the 18 different experiments mentioned earlier and elaborated in the Appendix.

\section{Results:}
First in Section \ref{sec:closedform}, we discuss the ability to apply our methods to different nonlinear kinetic rate law formulations within the context of the gluconeogenic model. %Then, in Section we discuss the different experimental design strategies that were found through our identifiability analysis to enable kinetic parameter estimation for the each flux in the model using steady state data. Finally, in Section we expand the application of our method to a relatively large (39 reactions) kinetic model of the red blood cell metabolism.
\subsection{Getting closed-form expressions for each flux in Kotte model:}\label{sec:closedform}
As demonstrated for identifying parameters of $v_1$ earlier in Section \ref{sec:example}, the computational effort required to establish identifiability is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. Although computer algebra systems are capable of handling complex symbolic calculations, sometimes, the complexity of getting closed form expressions for all kinetic parameters of certain rate law formulations is too much for the computer algebra system to overcome. We encounter this scenario when expression for the the allosterically activated reaction $v_3$ is described by the MWC model for allosteric regulation. We found that the complexity of the MWC kinetic rate law precluded its use to determine closed form expressions for the parameters in $v_3$ using either Mathematica or SymPy in Python. In order to overcome this computational obstacle, we modeled the reaction rate for $v_3$ using the convenience kinetic rate law formulation (reference for convenience kinetics). The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}). 
\begin{align}\label{eq:flux3_convkin}
content...
\end{align}

Using this expression for identifiability analysis, we found that each of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ have two different close form expressions owing to the presence of a square root term in their solutions. These distinct expressions are denoted by (1) and (2) following the respective parameter names throughout the rest of the document.

We believe this problem of computational tractability to occur in other complex kinetic rate law formulations as well. Example?

\subsection{Variability in the degree of identifiability of each parameter:}
We found that the ability of a combination of experiments to identify a particular parameter widely varies depending the the position of the flux in the metabolic network with respect to the inputs and the outputs of the network. In Figure \ref{fig:figure1} we show the number of combinations of experiments that are capable of identifying each parameter in each flux. In the left panels (Figure \ref{fig:figure1}a-c), the data is show when noise-free data is used for identification. The identifiability information for parameters using data with additive noise is shown on the right panel (Figure \ref{fig:figure1}d-f).

Both in the presence as well as absence of noise, the identifiability of parameters in $v_1$ is guaranteed by only about 20\% of the available data combinations (Figure \ref{fig:figure1}a, d). As seen in Section \ref{sec:example}, the identifiability expressions for both parameters in $v_1$ are dependent on changes in the both the uptake flux $v_1$ as well as the acetate concentration $ac$ in both the experiments used for identification. If either of these values cannot be distinguished between the two experiments whose data is used for identification, then the corresponding combination of experiments can be guaranteed to result in practical non-identifiability of the parameters in $v_1$. This is further illustrated in Figure below where the different experiments in each combination are shown.

When the experimental data has no measurement noise associated with it, then $V_2^{max}$ is the only parameter that can be identified by the most number of experimental combinations (89.5\%). In contrast to the high identifiability of this parameter, $K_2^{pep}$, the other parameter for $v_2$ is identified by less than 2\% of the available data combinations.

This is followed by $V_3^{max}$ that is capable of being identified by 76.3\% of the available data combinations.

In the small model that we use to elucidate our methodology, we find that only 
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/noise_no_noise_kotte_ident_jan01}}
	\caption{The number of data sets, made from a combination of 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. In the presence of additive noise in the flux and concentration data used for identifiability, the number of data sets practically identifying parameters in fluxes d) $v_1$, e) $v_2$ and f) $v_3$ are shown. 25 samples of additive noise $\epsilon ~ N(0, 0.05)$ were used to generate noisy data for figures in the right hand side panel.}\label{fig:figure1}
\end{figure}

\subsection{Utility of data for identifiability purposes:}

@@

\subsection{Combinations of experiments that will enable identification of all model parameters:}

\subsection{Expanding methodology to RBC model:}
If any data generated from a perturbation experiment $i$ results in nonidentifiability, we eliminate experiment $i$ from the list of experiments that need to be performed for parameter estimation.

\section*{Appendix}
\subsection*{In silico perturbation experiments for gluconeogenic model:}
\begin{table}[!tbhp]
	\caption{Table showing the perturbed values of all fluxes used for parameter estimation.}
	\begin{center}				
		\begin{tabular}{ccc}
			Designation & Perturbed Fluxes & Perturbed Values\\
			\hline
			P1 & $v_1$ & 2\\
			P2 & $v_2$ & 0.2\\
			P3 & $v_3$ & 0.5
		\end{tabular}
	\end{center}	
	\label{tab:pval}
\end{table}
